---
title: "WWF Canada Goose Report"
author: By "Xinzhou Du"
date: Published "11/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## 1. Introduction

Canada Goose *Branta canadensis* is one of the most populous waterfowl species in
North America.  Previously, the species is known to migrate from their breeding grounds
in Canada to their winter ranges in parts of the United States.  However, with conditions
in many of their breeding habitat becoming more temperate in Canada, food source are 
less scarce and consequently, less populations are migrating.  With local populations
projected to increase, it is crucial for the World Wide Fund (WWF) as a conservation
organization to understand the population trend over time and across landscapes to
better estimate corresponding environmental impacts.  In summary, this report projects
the population trend of the Canada Goose in Canada and attempts to attribute such trends
to various factors.

For access to modeling code, please refer to `goose.R`.  The R script contains code
used to clean and manipulate data, visualize data under various factors, statistical
analysis, and model construction.


### 1.1. Data Source

```{r, include = FALSE}
source("script/explore.R", local = knitr::knit_global())
```

We utilize population data from [The Living Planet Index (LPI)](https://livingplanetindex.org/home/index) to construct a mixed effect model
estimating how population of Canada Goose changes over years and across biomes.
The LPI database aggregates observation of species populations from a large subset
of studies around the world.  For Canada Goose in Canada, data are sourced from the
following studies.

```{r, echo=FALSE}
goose %>% 
        group_by(Data.source.citation) %>%
        summarize(count = n()) # most from Environment Canada, 471 obs
```


All data used are open source and are available for use for any registered users.


### 1.2. Research Question

Initially, we aim to investigate how does the abundance of Canada Goose varies across `years` 
and `biome` while taking into account random effects from location of monitoring, monitoring methods, and units defining abundance.  However, we have deviated from such question due to
lack of levels on certain variables and redundancies that we will dicuss later.

Our reformed research question looks at the same trend by only include the random effect
of the **locations of observations (provinces of Canada)**.


### 1.3. Hypothesis

Given the improved breeding ground conditions in Canada during winter, we expect 
the abundance of Canada Goose to remain similar over time but to differ across 
biome.


## 2. Exploratory Analysis

### 2.1. Distribution of Population Abundance

The original data distribution is difficult to interpret in aggregate as the sampling
methods differ, resulting in different units of observations.  We then attempted to
scale the data via standardization (scaling by mean and standard deviation).

We then plotted the resulting distribution.  At a glance, the data appears to be
heavily right skewed.

```{r, echo=FALSE}
ggplot(data = goose, aes(x = sscalepop)) +
        geom_histogram(bins = sqrt(nrow(goose))) + 
        style() +
        xlab("Standardized Population") +
        ylab("Frequency")
```

This is further reflected in the scatter plot below with scaled popualtion over time.
We can see there is a positive exponential relationship between year and standardized
population.

```{r, echo=FALSE}
ggplot(data = goose, aes(x = year, y = sscalepop)) +
        geom_point() + 
        style() +
        xlab("Year") +
        ylab("Standardized Population")
```


This suggests we should log scale our original population data.  We then investigated 
the log scaled data divided by different location of observation.  We can see that
observations conducted in Southern James Bay, a region within Ontario, has disproportionately
larger observed population in earlier years than other data points.

```{r, echo=FALSE}
ggplot(data = goose, aes(x = year, y = logpop, color = Location.of.population)) +
        geom_point() + 
        style() +
        labs(title = "Goose Population Over Time Across Location of Observations", x = "Year", y = "log(Population)", color = "Locations") 
```


With a box plot, we can also see how at Southern James Bay and Nestor One study Area, different methods of abundance survey were utilized, which possibly resulted in their different trends.
We therefore did not consider the two sites for our model analysis.

```{r, echo=FALSE}
ggplot(data = goose, aes(x = Location.of.population, y = logpop, color = Sampling.method)) +
        geom_boxplot() + 
        style() +
        theme(axis.text.x = element_text(angle = 90)) 
```


Another complication presented towards our analysis is the relationship between biome and location of our observation.  We can see from the boxplot below that  one location
of observation contains exactly one biome, but some biomes are more present across locations
than others, notably boreal and taiga forest.

```{r, include = FALSE}
source("script/goose.R", local = knitr::knit_global())
```

```{r, echo=FALSE}
ggplot(data = goose, aes(x = Location.of.population, y = logpop, color = biome)) +
        geom_boxplot() + 
        style() +
        theme(axis.text.x = element_text(angle = 90))
```


### 2.2. Preliminary Statistical Analysis

We conducted ANOVAs between our 3 variables of interest: `year`,`biomes`, `Location.of.observation`.

```{r, echo=FALSE}
stargazer(lm(logpop ~ I(year-1969), data = goose), type = "text",
          digits = 3,
          star.cutoffs = c(0.05, 0.01, 0.001),
          digit.separator = "")

stargazer(lm(logpop ~ biome, data = goose), type = "text",
          digits = 3,
          star.cutoffs = c(0.05, 0.01, 0.001),
          digit.separator = "")

stargazer(lm(logpop ~ Location.of.population, data = goose), type = "text",
          digits = 3,
          star.cutoffs = c(0.05, 0.01, 0.001),
          digit.separator = "")
```

In summary, we found our log scaled population to differ significantly across levels 
of all 3 factors.  We found that a linear model with `year` explaining variations in
log transformed population abundance to retain an adjusted R-squared value of 0.063,
suggested a high amount of unexplained residuals.

In addition, we only have 4 levels for biome, so we have to set it as a fixed effect.


## 3. Model Analysis

### 3.1. Model Building

We first construct our base models, which attempts to explain variation in log transformed
population abundance by `year`, `biome`, and `Location.of.population` individually.


**Base Model 1 (bm1)**

$$
log(Population) \sim year
$$


**Base Model 2 (bm2)**

$$
log(Population) \sim biome
$$


**Base Model 3 (bm3)**

$$
log(Population) \sim location
$$

We then constructed a mixed effect model utilizing all 3 variables. Because we are primarily interested in how Canada Goose population changes over time and if populations changes also 
differ across biomes, we set `year` and `biome` as fixed effects and `Location.of.population`
as a random effect.


**Mixed Model 1**

$$
log(Population) \sim year + biome + (1|location)
$$

We also attempted to remove biome as a fixed effect to compare model fitness against
our mixed model above.


**Mixed Model 2**

$$
log(Population) \sim year + (1|location)
$$

We compared these models using the Akaike information criterion (AIC) to determine the 
parsimonious with respect to our data.  In R, we utilized the AICc as the information 
criterion because it invokes a heavier penalty for each additional parameter estimated,
preventing us from "overfitting" our model. 


```{r, echo=FALSE}
AICc(bm1, bm2, bm3, mm1, mm2)
```

We found our model utilizing `year` and `biome` as fixed effect resulted in the lowest 
AICc (AICc = 572.54) compared to the other model, despite having the second highest degree of freedom (df = 7).  This indicates out of our 5 proposed models, mixed model 1 `mm1` retains 
the highest model likelihood and has substantial explanatory power in predicting population
abundance of Canada Goose.  We therefore select `mm1` as our model for visualization and diagnostics.


### 3.2. Model analytics

Below is the summary output of our selected Mixed Model.

```{r, echo=FALSE}
summary(mm1)
```

Here we see the variance for our only random effects group `Location.of.population` to be non-zero (Variance = 0.1068), which suggest that it is in fact necessary to include this into our model.  The variance attributed to our random effect group is also on par with the residual variance (Variance = 0.1774) which indicates indicates population abundances across different locations are indeed different from each other.

In the correlation of fixed effect section, we see that none of our explanatory values have correlation values greater than 0.3 with each other.  This indicating the lack of covariance across our predictor variables allowing us to conduct meaningful statistical inferences.

To test the significance of our fixed effect, we compare our selected model a null model with a value of 1 for our 2 fixed effects.  We then compared them based on their resulting AICc.


**Null Model (null)**

$$

log(Population) \sim 1 + 1 + (1|location)

$$

```{r, echo=FALSE}
AICc(null, mm1)
```
From here, we see that our selected model with `year` and `biome` as fixed effects have much lower AICc compared to the null mode without any fixed effects.  This suggests the significance of including our selected fixed effects within our model.

From these above analyses, we see including our selected fixed and random effects improve our ability to explain changes in Canada Goose population. 

### 3.3. Model Visualization

The first plot visualizes a fitted line using our selected model and the original log transformed population trend of Canada goose over the years and how they differ by biomes.

```{r, echo=FALSE}
ggplot(pred.mm) +
        geom_line(aes(x = (x+1969), y = predicted)) +
        geom_ribbon(aes(x = (x+1969), ymin = predicted - std.error, ymax = predicted + std.error), 
                    fill = "lightgrey", alpha = 0.5) + 
        geom_point(data = goose, aes(x = year, y = logpop, color = biome)) +
        labs(x = "Year", y = "log(population)", 
             title = "Change in Canada Goose Population Over Time") +
        style()
```

We can see our fitted line have relatively small error margin and seem to predict population trends in the Boreal and Taiga biome way better than the other biomes.  This is also possibily influenced by the large amount of data observations conducted in this biome.

We then illustrate the influence our each levels of `Location.of.population` on our model output.

```{r, echo=FALSE}
plot_model(mm1, type = "re", show.values = TRUE)
```

We can see data points from from Ontario, Nova Scotia, Northwest Territories, and New Brunswick deviates the most from our model output.  Data points from Yukon deviates the least.


## 4. Model Diagnostic

As previously stated, it is highly likely that our selected model is better at predicting certain values of population abundances compared to others.  We illustrate this by utilizing a residual plot.

```{r, echo=FALSE}
plot(mm1)
```

We see the residual values are negative for lower fitted value of population abundances and positive for fitted values around -1 to 1.  Revisiting our model visualization, it might be more useful in the future to instead model different biomes independently since their population trends appears to have significantly different y-intercepts and slopes.  This is especially the case when comparing data points from temporate mixed forests and boreal and taiga forests.


## 5. Conclusion

Throughout this report, we presented a mixed linear model predicting the population trend of Canada Goose in Canada.  We utilized year of observation and biome of the goose population as fixed effect and the location of observation as random effect.  We demonstrated through a model selection process under AICc that our model is significant in its explanatory value.  

Because we showed that year and biome as fixed effects are significant explanatory factors for changes in populationo abundance, we reject our initially registered hypothesis that population abundance do not differ overtime.

